---
title: "DataPreparation"
author: "Edoardo Diana"
date: "2025-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
fig.width = 10, fig.height = 7)
```


# Introduction

This notebook prepares and explores the **arXiv General Relativity collaboration network** dataset.

## Dataset Description

- **Source**: Stanford Network Analysis Project (SNAP)
- **Period**: 1993-2003
- **Domain**: General Relativity and Quantum Cosmology papers
- **Format**: Undirected graph (co-authorship)

---

# Load Libraries and Data

```{r}
library(igraph)
library(dplyr)
library(ggplot2)
library(knitr)

source("../scripts/00_setup.R")



#Scarica dati se non presenti (esegue 01_download_data.R)
source("../scripts/01_download_data.R")

# Load network
g <- readRDS("data/processed/network.rds")




```

---

# Basic Network Properties


```{r}
cat("Nodes:", vcount(g), "\n")
cat("Edges:", ecount(g), "\n")
cat("Density:", edge_density(g), "\n")
cat("Diameter:", diameter(g), "\n")
cat("Average path length:", mean_distance(g), "\n")
cat("Connected:", is_connected(g), "\n")

```

---

# Degree Distribution


```{r}

deg <- degree(g)

# Summary statistics
summary(deg)

# Plot
ggplot(data.frame(degree = deg), aes(x = degree)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  labs(title = "Degree Distribution",
  x = "Degree", y = "Frequency") +
  theme_minimal()
```

---

# Network Visualization (Subset)


```{r}

# Take largest connected component for visualization
comp <- components(g)
g_main <- induced_subgraph(g, which(comp$membership == which.max(comp$csize)))

# Sample 500 nodes for readability
set.seed(42)
g_sample <- induced_subgraph(g_main, sample(vcount(g_main), 1000))

#Plot
plot(g_sample,
  vertex.size = 2,
  vertex.label = NA,
  vertex.color = "lightblue",
  edge.color = "gray80",
  layout = layout_with_fr(g_sample),
  main = "Scientific Collaboration Network (Sample)")
```

---

# Save Prepared Data

Save main component for further analysis

```{r}
saveRDS(g_main, "data/processed/network_main_component.rds")
cat("✓ Main component saved\n")

```


---

# Next Steps

- **02_centrality_analysis.Rmd**: Identify influential researchers
- **03_community_detection.Rmd**: Detect research communities
- And so on...

```{r}


```

