---
title: "02_centrality_analysis"
author: "Edoardo Diana"
date: "2025-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
fig.width = 12, fig.height = 8)

```

```{r}
library(igraph)
library(dplyr)
library(ggplot2)
library(tidyr)
library(tidygraph)
library(ggraph)

```


# Introduction

This notebook analyzes **centrality measures** to identify the most influential researchers in the arXiv General Relativity collaboration network.

## Research Question

**Who are the most influential researchers in the theoretical physics community?**

Different centrality measures capture different notions of "influence":

- **Degree**: Social hubs (many collaborators)
- **Closeness**: Quick access to entire network
- **Betweenness**: Knowledge brokers (bridge different communities)
- **Eigenvector**: Collaborate with other influential researchers
- **PageRank**: Prestige from connections
- **Katz**: Weighted influence considering all path lengths

---

# Load Data

Load main component (prepared in 01_data_preparation.Rmd)

```{r}
data_file <- "data/processed/network_main_component.rds"

if (!file.exists(data_file)) {
stop("❌ Main component not found. Run 01_data_preparation.Rmd first!")
}

g <- readRDS(data_file)

cat("Network loaded:\n")
cat("Nodes:", vcount(g), "\n")
cat("Edges:", ecount(g), "\n")
cat("Density:", round(edge_density(g), 6), "\n")

```

---
## Power Function

Compute power \( x = (1/x) A \) using iterative method.

```{r}
# COPIATA DAL PROFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFF

# Compute power x = (1/x) A 
#INPUT
# A = graph adjacency matrix
# t = precision
# OUTPUT
# A list with:
# vector = power vector
# iter = number of iterations

power = function(A, t) {
  n = dim(A)[1];
  # x_2k
  x0 = rep(0, n);
  # x_2k+1
  x1 = rep(1, n);
  # x_2k+2
  x2 = rep(1, n);
  diff = 1
  eps = 1/10^t;
  iter = 0;
  while (diff > eps) {
    x0 = x1;
    x1 = x2;
    x2 = (1/x2) %*% A;
    diff = sum(abs(x2 - x0));
    iter = iter + 1;
  } 
  # it holds now: alpha x2 = (1/x2) A
  alpha = ((1/x2) %*% A[,1]) / x2[1];
  # hence sqrt(alpha) * x2 = (1/(sqrt(alpha) * x2)) A
  x2 = sqrt(alpha) %*% x2;
  return(list(vector = as.vector(x2), iter = iter))
}  
```


# Centrality Measures


---

# Calculate All Centrality Measures



```{r}

# Extract adjacency matrix
A <- as_adjacency_matrix(g, sparse = TRUE)

# 1. DEGREE CENTRALITY
V(g)$degree <- degree(g)

# 2. CLOSENESS CENTRALITY
V(g)$closeness <- closeness(g, normalized = TRUE)

# 3. BETWEENNESS CENTRALITY
V(g)$betweenness <- betweenness(g, normalized = TRUE)

# 4. EIGENVECTOR CENTRALITY
V(g)$eigenvector <- eigen_centrality(g)$vector

# 5. PAGERANK
V(g)$pagerank <- page_rank(g)$vector

# 6. KATZ CENTRALITY (alpha based on eigenvalues, like professor)

# Metodo troppo lento
# eig <- eigen(A)$values
# r <- max(abs(eig))
# alpha <- 0.85 / r
# V(g)$katz <- alpha_centrality(g, alpha = alpha)

eig_result <- eigen_centrality(g)
lambda_max <- eig_result$value # Dominant eigenvalue from igraph
alpha <- 0.85 / lambda_max
V(g)$katz <- alpha_centrality(g, alpha = alpha)
cat("✓ Katz computed (alpha =", round(alpha, 6), ", lambda_max =", round(lambda_max, 2), ")\n")


cat("✓ All centrality measures computed and saved as vertex attributes\n")

# 7. POWER

# Convert sparse matrix to dense (required for power function)
A_dense <- as.matrix(A)

# Add damping to adjacency matrix
damping <- 0.15
n <- vcount(g)
I <- diag(damping, n)
Ad <- A_dense + I

# Compute power with precision t = 4 (faster than t = 6)
cat("Computing power... (this may take several minutes)\n")
power_result <- power(Ad, t = 4)
V(g)$power <- power_result$vector

cat("✓ Power computed (", power_result$iter, "iterations)\n")


cat("\n✓ All centrality measures and power computed and saved as vertex attributes\n")
```


---

# Extract Node Data

Now extract vertex attributes into a dataframe (professor's style with `as_data_frame`).

```{r}
# Extract all vertex attributes as dataframe
nodes <- as_data_frame(g, what = "vertices")

# Add node_id if not present
if (!"name" %in% colnames(nodes)) {
  nodes$name <- as.character(1:vcount(g))
}

# Add normalized versions
nodes <- nodes %>%
  mutate(
    degree_norm = (degree - min(degree)) / (max(degree) - min(degree)),
    closeness_norm = (closeness - min(closeness)) / (max(closeness) - min(closeness)),
    betweenness_norm = (betweenness - min(betweenness)) / (max(betweenness) - min(betweenness)),
    eigenvector_norm = (eigenvector - min(eigenvector)) / (max(eigenvector) - min(eigenvector)),
    pagerank_norm = (pagerank - min(pagerank)) / (max(pagerank) - min(pagerank)),
    katz_norm = (katz - min(katz)) / (max(katz) - min(katz))
  )

cat("✓ Node dataframe created with", nrow(nodes), "researchers\n")
head(nodes)
```


```{r}
```

