---
title: "02_centrality_analysis"
author: "Edoardo Diana"
date: "2025-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
fig.width = 12, fig.height = 8)

```

```{r}
library(igraph)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(patchwork) # Per combinare plot

```


# Introduction

This notebook analyzes **centrality measures** to identify the most influential researchers in the arXiv General Relativity collaboration network.

## Research Question

**Who are the most influential researchers in the theoretical physics community?**

Different centrality measures capture different notions of "influence":

- **Degree**: Social hubs (many collaborators)
- **Closeness**: Quick access to entire network
- **Betweenness**: Knowledge brokers (bridge different communities)
- **Eigenvector**: Collaborate with other influential researchers
- **PageRank**: Prestige from connections
- **Katz**: Weighted influence considering all path lengths

---

# Load Data

Load main component (prepared in 01_data_preparation.Rmd)

```{r}
data_file <- "data/processed/network_main_component.rds"

if (!file.exists(data_file)) {
stop("❌ Main component not found. Run 01_data_preparation.Rmd first!")
}

g <- readRDS(data_file)

cat("Network loaded:\n")
cat("Nodes:", vcount(g), "\n")
cat("Edges:", ecount(g), "\n")
cat("Density:", round(edge_density(g), 6), "\n")

```

---

# Centrality Measures


---

# Calculate All Centrality Measures



```{r}
# Calculate all centrality measures
centrality_df <- data.frame(
  node_id = as.character(1:vcount(g)),
  degree = degree(g),
  closeness = closeness(g, normalized = TRUE),
  betweenness = betweenness(g, normalized = TRUE),
  eigenvector = eigen_centrality(g)$vector,
  pagerank = page_rank(g)$vector,
  katz = alpha_centrality(g, alpha = 1/(max(eigen(as_adjacency_matrix(g, sparse = FALSE))$values) + 0.1))
)

# Add normalized versions for comparison​
centrality_df <- centrality_df %>%
  mutate(
    degree_norm = (degree - min(degree)) / (max(degree) - min(degree)),
    closeness_norm = (closeness - min(closeness)) / (max(closeness) - min(closeness)),
    betweenness_norm = (betweenness - min(betweenness)) / (max(betweenness) - min(betweenness)),
    eigenvector_norm = (eigenvector - min(eigenvector)) / (max(eigenvector) - min(eigenvector)),
    pagerank_norm = (pagerank - min(pagerank)) / (max(pagerank) - min(pagerank)),
    katz_norm = (katz - min(katz)) / (max(katz) - min(katz))
  )

cat("✓ Centrality measures calculated\n")

```


---

# 1. Degree Centrality

**Definition**: Number of direct collaborators (social hubs).


```{r}

# Top 15 by degree
top15_degree <- centrality_df %>%
  arrange(desc(degree)) %>%
  head(15) %>%
  mutate(node_id = factor(node_id, levels = node_id))

# Barplot
ggplot(top15_degree, aes(x = node_id, y = degree, fill = degree)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Top 15 Researchers by Degree Centrality",
    subtitle = "Number of direct collaborators",
    x = "Author ID", y = "Degree (Number of Collaborators)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
  axis.text.y = element_text(size = 10))
```


---

# 2. Closeness Centrality

**Definition**: Average distance to all other nodes (quick access to network).


```{r}
top15_closeness <- centrality_df %>%
  arrange(desc(closeness)) %>%
  head(15) %>%
  mutate(node_id = factor(node_id, levels = node_id))

ggplot(top15_closeness, aes(x = node_id, y = closeness, fill = closeness)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightyellow", high = "orange") +
  labs(title = "Top 15 Researchers by Closeness Centrality",
    subtitle = "Quick access to entire network",
    x = "Author ID", y = "Closeness (normalized)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

```

---

# 3. Betweenness Centrality

**Definition**: Bridges between communities (knowledge brokers).



```{r}
top15_betweenness <- centrality_df %>%
  arrange(desc(betweenness)) %>%
  head(15) %>%
  mutate(node_id = factor(node_id, levels = node_id))

ggplot(top15_betweenness, aes(x = node_id, y = betweenness, fill = betweenness)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen") +
  labs(title = "Top 15 Researchers by Betweenness Centrality",
    subtitle = "Knowledge brokers connecting communities",
    x = "Author ID", y = "Betweenness (normalized)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

```


---

# 4. Eigenvector Centrality

**Definition**: Influence from connections to other influential nodes (academic prestige).


```{r}
top15_eigen <- centrality_df %>%
  arrange(desc(eigenvector)) %>%
  head(15) %>%
  mutate(node_id = factor(node_id, levels = node_id))

ggplot(top15_eigen, aes(x = node_id, y = eigenvector, fill = eigenvector)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "pink", high = "red") +
  labs(title = "Top 15 Researchers by Eigenvector Centrality",
    subtitle = "Collaborating with other influential researchers",
    x = "Author ID", y = "Eigenvector Centrality") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

```


---

# 5. PageRank

**Definition**: Google's algorithm for network prestige.


```{r}

top15_pagerank <- centrality_df %>%
  arrange(desc(pagerank)) %>%
  head(15) %>%
  mutate(node_id = factor(node_id, levels = node_id))

ggplot(top15_pagerank, aes(x = node_id, y = pagerank, fill = pagerank)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightcoral", high = "darkred") +
  labs(title = "Top 15 Researchers by PageRank",
    subtitle = "Overall network prestige",
    x = "Author ID", y = "PageRank") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

---

# 6. Katz Centrality

**Definition**: Weighted influence considering all path lengths.



```{r}

top15_katz <- centrality_df %>%
  arrange(desc(katz)) %>%
  head(15) %>%
  mutate(node_id = factor(node_id, levels = node_id))

ggplot(top15_katz, aes(x = node_id, y = katz, fill = katz)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lavender", high = "purple") +
  labs(title = "Top 15 Researchers by Katz Centrality",
    subtitle = "Global influence via all network paths",
    x = "Author ID", y = "Katz Centrality") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```


---

# Comparative Analysis

## Distribution of All Centrality Measures



```{r}

plot_data <- centrality_df %>%
  select(degree, closeness, betweenness, eigenvector, pagerank, katz) %>%
  pivot_longer(everything(), names_to = "measure", values_to = "value")

ggplot(plot_data, aes(x = value, fill = measure)) +
  geom_histogram(bins = 50, alpha = 0.7, show.legend = FALSE) +
  facet_wrap(~ measure, scales = "free", ncol = 2) +
  labs(title = "Distribution of Centrality Measures",
    x = "Centrality Value", y = "Frequency") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold", size = 12))
```

---

## Top 10 Comparison (Side-by-side)


```{r}

# Get top 10 by PageRank
top10_ids <- centrality_df %>%
  arrange(desc(pagerank)) %>%
  head(10) %>%
  pull(node_id)

# Extract normalized values for comparison
top10_data <- centrality_df %>%
  filter(node_id %in% top10_ids) %>%
  select(node_id, degree_norm, closeness_norm, betweenness_norm,
  eigenvector_norm, pagerank_norm, katz_norm) %>%
  pivot_longer(-node_id, names_to = "measure", values_to = "value") %>%
  mutate(measure = gsub("_norm", "", measure),
    measure = factor(measure, levels = c("degree", "closeness", "betweenness",
"eigenvector", "pagerank", "katz")))

ggplot(top10_data, aes(x = node_id, y = value, fill = measure)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  coord_flip() +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Top 10 Researchers: Normalized Centrality Comparison",
       subtitle = "All measures scaled to [0,1] for fair comparison",
       x = "Author ID", y = "Normalized Centrality",
       fill = "Measure") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

```


---

## Correlation Heatmap



```{r}
# Correlation matrix
cor_matrix <- cor(centrality_df[, c("degree", "closeness", "betweenness",
"eigenvector", "pagerank", "katz")])

# Convert to long format for ggplot
cor_long <- as.data.frame(as.table(cor_matrix))
  names(cor_long) <- c("Var1", "Var2", "Correlation")

ggplot(cor_long, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Correlation, 2)), color = "black", size = 4) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
  midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Correlation Matrix of Centrality Measures",
    x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


---

# Network Visualizations with ggraph

## Network Colored by Degree


```{r}
# Sample nodes for visualization
set.seed(42)

# Get all node IDs from centrality_df (consistent source)
all_node_ids <- centrality_df$node_id

# Top 200 by degree
top_nodes <- centrality_df %>%
  arrange(desc(degree)) %>%
  head(200) %>%
  pull(node_id)

# Remaining nodes
remaining_nodes <- setdiff(all_node_ids, top_nodes)

# Sample 300 from remaining (or all if < 300)
n_sample <- min(300, length(remaining_nodes))
random_nodes <- sample(remaining_nodes, n_sample)

# Combine
sample_nodes <- c(top_nodes, random_nodes)

cat("Sample size:", length(sample_nodes), "nodes\n")
cat(" - Top nodes:", length(top_nodes), "\n")
cat(" - Random nodes:", n_sample, "\n")

# Create subgraph
g_sample <- induced_subgraph(g, sample_nodes)

# Convert to tidygraph
tg_sample <- as_tbl_graph(g_sample) %>%
  activate(nodes) %>%
  mutate(degree_cent = centrality_degree())

# Plot with ggraph
ggraph(tg_sample, layout = "fr") +
  geom_edge_link(alpha = 0.1, color = "gray70") +
  geom_node_point(aes(size = degree_cent, color = degree_cent)) +
  scale_color_gradient(low = "lightblue", high = "darkblue", name = "Degree") +
  scale_size_continuous(range = c(1, 8), guide = "none") +
  labs(title = "Collaboration Network Colored by Degree Centrality",
    subtitle = "Node size and color represent degree (number of collaborators)") +
  theme_graph() +
  theme(legend.position = "right")

```


---

## Network Colored by Betweenness



```{r}

tg_sample <- tg_sample %>%
  activate(nodes) %>%
  mutate(betw_cent = centrality_betweenness(normalized = TRUE))

ggraph(tg_sample, layout = "fr") +
  geom_edge_link(alpha = 0.1, color = "gray70") +
  geom_node_point(aes(size = betw_cent, color = betw_cent)) +
  scale_color_gradient(low = "lightgreen", high = "darkgreen", name = "Betweenness") +
  scale_size_continuous(range = c(1, 8), guide = "none") +
  labs(title = "Collaboration Network Colored by Betweenness Centrality",
    subtitle = "Larger nodes = knowledge brokers connecting communities") +
  theme_graph() +
  theme(legend.position = "right")
```

---

## Network Colored by PageRank


```{r}
tg_sample <- tg_sample %>%
  activate(nodes) %>%
  mutate(pr_cent = centrality_pagerank())

ggraph(tg_sample, layout = "fr") +
  geom_edge_link(alpha = 0.1, color = "gray70") +
  geom_node_point(aes(size = pr_cent, color = pr_cent)) +
  scale_color_gradient(low = "lightyellow", high = "red", name = "PageRank") +
  scale_size_continuous(range = c(1, 8), guide = "none") +
  labs(title = "Collaboration Network Colored by PageRank",
    subtitle = "Red nodes = High prestige researchers") +
  theme_graph() +
  theme(legend.position = "right")

```

---

# Key Findings

## Summary: Top Researchers by Measure


```{r}
# Identify top 5 for each measure
top5_all <- bind_rows(
  centrality_df %>% arrange(desc(degree)) %>% head(5) %>% mutate(measure = "Degree"),
  centrality_df %>% arrange(desc(closeness)) %>% head(5) %>% mutate(measure = "Closeness"),
  centrality_df %>% arrange(desc(betweenness)) %>% head(5) %>% mutate(measure = "Betweenness"),
  centrality_df %>% arrange(desc(eigenvector)) %>% head(5) %>% mutate(measure = "Eigenvector"),
  centrality_df %>% arrange(desc(pagerank)) %>% head(5) %>% mutate(measure = "PageRank"),
  centrality_df %>% arrange(desc(katz)) %>% head(5) %>% mutate(measure = "Katz")
) %>%
  select(node_id, measure, degree_norm, betweenness_norm, pagerank_norm) %>%
  pivot_longer(c(degree_norm, betweenness_norm, pagerank_norm),
    names_to = "metric", values_to = "value")

ggplot(top5_all, aes(x = node_id, y = value, fill = metric)) +
  geom_col(position = "dodge") +
  facet_wrap(~ measure, scales = "free_x", ncol = 2) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2", labels = c("Betweenness", "Degree", "PageRank")) +
  labs(title = "Top 5 Researchers per Centrality Measure",
    subtitle = "Comparing their profiles across different metrics",
    x = NULL, y = "Normalized Centrality",
    fill = "Metric") +
  theme_minimal() +
  theme(legend.position = "bottom",
  axis.text.y = element_text(size = 8))

```

---

# Save Results


```{r}
# Save centrality data
write.csv(centrality_df, "output/tables/centrality_all_measures.csv", row.names = FALSE)

cat("✓ Results saved to output/tables/centrality_all_measures.csv\n")

```


---

# Conclusions

1. **Degree centrality** identifies social hubs with many collaborators
2. **Betweenness centrality** reveals knowledge brokers connecting research groups
3. **PageRank and Eigenvector** highlight academic prestige
4. Different measures capture **complementary dimensions** of influence

**Next**: Community detection to see if top-central nodes are also community leaders.

```{r}


```